Raft Distributed Consensus Algorithm over\\ WebRTC (Web Real-Time Communication)

### 2015-12-11 research_class/paper.tex
paper.tex: trimed to 10 pages without OPTIONALS.

In order to keep the replicated transaction log from growing
indefinitely, Raft nodes should periodically compact their logs. There
are many different ways to accomplish log compaction and the best way
will depend on application requirements and on the specific nature of
the distributed state machine. For example, if the state machine is
simply a single shared counter, then any log entries before the most
recently committed entry may be safely discarded. The simplest and
most generic solution is often to provide some way of snapshotting the
entire state machine (including the term and log index it represents)
to disk at which point the all previously applied log entries can be
discarded. Snapshotting the state machine also means that the
implementation must provide a way for the leader to serialize and send
the current state machine to other nodes if their transaction log is
too old (e.g. when a new node is added)
\cite[Section~5]{raft_thesis:ongaro14}.

@@

The Raft protocol can also provide strict linearizable semantics for
client commands/requests (reads and writes) for applications that
require it. In order to accomplish this, each client is given a unique
ID, and the client assigns a unique serial number to each command.
This prevents a single client command from being applied twice (leader
crash, network duplication) or lost (packet loss)
\cite[Section~6.3]{raft_thesis:ongaro14}.

@@

\begin{figure*}[!t]
\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_architecture}}
\caption{Raft over RTC Architecture}
  \label{fig:raft_rtc_architecture}
\end{figure*}

\subsection{Raft.js}

The modular design of the Raft.js implementation allows it to be
easily used in different contexts. For example, the \texttt{RaftServerLocal}
class (\texttt{local.js}) implements RPC calls using plain JavaScript function
calls between different instances of the class (nodes) in the same
execution context. For example, the online Raft visualization at
\href{http://kanaka.github.io/raft.js/}{http://kanaka.github.io/raft.js/}
\cite{raft.js:visualization} uses the \texttt{RaftServerLocal} class
and instantiates it using a scheduling function that executes one
scheduled event each time the user clicks on the ``Step" button rather
than due to the passage of time.

The current version of Raft.js implements most of the Raft algorithm
including: leader election, log replication, safety, membership
changes, basic client interaction, and read-only operation
optimizations. % TODO: read-only operations needs mentioning earlier
Log compaction and full client linearizability are not currently
implemented (clients and client transactions are not assigned unique
IDs). Log compaction is important for long-running production systems
in order to limit memory growth and allow new nodes to catch up their
logs more efficiently. However log compaction is not required for
verifying overall system correctness and does not significantly impact
short-running test configurations. Full client linearizability is
an important feature for simplifying certain types of distributed
applications, however the implementation of linearizability requires
some significant additions to the Raft implementation (e.g. assigning
client IDs and sequence IDs and tracking timed-out clients). For the
chat application, simply adding a single atomic check-and-set state
machine command is sufficient to achieve a linearized chat history.
This does impose a slight additional cost on the client because it
must retry the atomic operations until they succeed. The author
intends to implement both log compaction and full client
linearizability in the future, however these do not have a significant
impact on the ability to test Raft combined with WebRTC.

\begin{enumerate}
\item The paper describes the Raft RPCs in terms of synchronous stream
    based request/response model where requests and responses are
    automatically correlated together by a lower level of the software
    stack. For example with HTTP 1.0 \cite{http:rfc1945} client
    requests are followed by a server response on any given TCP
    stream. Most remote procedure call mechanims also use
    a synchronous request/response model. %TODO/CITE
    WebRTC DataChannel connections are message based rather than
    request/response stream based. %TODO/CITE
    In the original Raft definition of RPC, request and response messages
    almost contain enough information to function correctly in an
    environment without builtin request/response correlation, such as
    UDP or SCTP when used with reliable transport disabled. %TODO/CITE
    Raft.js includes a small modification to the response RPCs for
    requestVote and appendEntries to add a `sourceId' field that
    enables the protocol to function with message based transports.
\item The Raft paper is clear that each membership change must be
    delayed until all the previous membership change is committed
    (sequential) \cite[Section~4.1]{raft_thesis:ongaro14}. It is
    implied that the leader node should track all pending membership
    change requests and apply them one at a time.  However, in
    Raft.js, if a client requests a membership change while another
    change request is pending (uncommitted), the leader will reject
    the change request with a status of `PENDING\_CONFIG\_CHANGE'.
    This keeps the core Raft implementation simpler with the trade-off
    being that the client implementation of membership change requests
    is slightly more complex due to retry logic.
\end{enumerate}

@@

\subsubsection{WebRTC Protocol}

@@

The protocols that make up the WebRTC suite are categorized into
five general functionality groups:

\begin{enumerate}
    \item data transport
        \ifdefined\OPTIONAL
        \\
        \cite{draft-rtcweb-transports}
        \cite{draft-rtcweb-qos}
        \cite{RFC4960:sctp}
        \cite{RFC3758:sctp-reliability}
        \cite{draft-sctp-dtls-encaps}
        \cite{draft-mmusic-sctp-sdp}
        \cite{draft-sctp-ndata}
        \cite{draft-rtcweb-rtp-usage}
        \cite{draft-rtcweb-data-channel}
        \cite{draft-rtcweb-data-protocol}
        \cite{draft-rtcweb-alpn}
        \cite{RFC5245:ice}
        \cite{RFC5128:p2p-nat}
        \cite{RFC5389:stun}
        \cite{RFC3489:stun-udp}
        \cite{RFC5766:turn}
        \cite{RFC6156:turn-ipv6}
        \cite{RFC6544:tcp-ice}
        \cite{RFC4571:rtp-rtcp}
        \cite{RFC5764:dtls-srtp}
        \fi
    \item data framing and security
        \ifdefined\OPTIONAL
        \\
        \cite{RFC3550:rtp}
        \cite{RFC3711:srtp}
        \cite{draft-rtcweb-rtp-usage}
        \cite{draft-rtcweb-data-channel}
        \cite{draft-rtcweb-data-protocol}
        \fi
    \item connection management
        \ifdefined\OPTIONAL
        \\
        \cite{RFC2327:sdp}
        \cite{RFC3264:sdp-offer-answer}
        \cite{draft-rtcweb-jsep}
        \cite{RFC5245:ice}
        \cite{draft-mmusic-trickle-ice}
        \cite{RFC5763:srtp-using-dtls}
        \fi
    \item data formats
    \item presentation and control
    \item local system support functions
\end{enumerate}

The last three functionality groups focus on interactive audio and video
applications and are less relevant to Raft over WebRTC. \ifdefined\OPTIONAL
Some of the important standards for the first three functionality
groups are cited above. \fi

@@

%\begin{figure*}[!t]
\begin{figure}[!t]
%\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_sequence}}
\centerline{\includegraphics[width=0.45\textwidth]{imgs/raft_rtc_sequence}}
\caption{Raft over RTC Sequence Diagram}
  \label{fig:raft_rtc_sequence}
\end{figure}

\section {Design and Implementation}

@@

The file `rtc\_server.js' implements the signaling server for the Raft
over WebRTC application. The server extends the PeerJS
`ExpressPeerServer' object in order to provide WebRTC signaling. In
addition to the signaling function, the server also provides the
following service endpoints:

@@

Subsequent nodes are added to the same Raft cluster by starting a new
browser context (separate browser, browser window, or browser tab) and
loading the same URL without the ``firstServer" fragment identifier
(Figure~\ref{fig:raft_rtc_sequence}~[9]). There is a convenience link
provided in the application page that opens a new browser window with
a new cluster node (see \ref{fig:browser_1}).

@@

In addition to the low level Raft message log, the following
statistics are shown in the application:

\begin{enumerate}
\item \emph{Term}: the current Raft leadership term.
\item \emph{State}: the current Raft node state (follower, candidate,
    or leader)
\item \emph{Cluster Size}: a count of the current Raft cluster
    membership (number of nodes currently participating in the
    cluster)
\item \emph{Log Length}: number of entries in the transaction log.
\item \emph{Request Votes Sent/Received}: the number of `requestVote`
    RPCs that have been sent from and received by this node.
\item \emph{Append Entries Sent/Received}: the number of
    `appendEntries' RPCs that have been sent from and received by this
    node.
\end{enumerate}

@@

In the Raft over WebRTC implementation each local browser is a client
of the Raft cluster. However in Raft, the leader node is the only node
that adds entries to the log so clients on follower nodes must have
a way to communicate with the leader. This is implemented as a pair of
functions named \texttt{clientRequest} and
\texttt{clientRequestResponse}. The \texttt{clientRequest} function is
called by the client with the arguments of a client request RPC and
a callback function to be called once the request is completed. The
\texttt{clientRequest}/\texttt{clientRequestResponse} functions work
as follows:

\begin{enumerate}
\item The \texttt{clientRequestResponse} is registered with the local
    Raft.js implementation as the callback function anytime
    a \texttt{clientRequestResponse} RPC is received.
\item The \texttt{clientRequest} function keeps track of the current
    leader ID.  If the current leader ID is not set or this node is
    the leader then the client request RPC is sent directly to the
    local Raft node bypassing the network transport layer.  If the
    current leader ID is set then the request is sent to that node as
    a normal RPC request over the WebRTC channel.
\item When the \texttt{clientRequestResponse} function is called and
    the status is ``NOT\_LEADER" then the current leader ID being
    tracked by the \texttt{clientRequest} function is set to the
    leader ID specified in the response. Then the original request is
    re-issued over the normal WebRTC channel to the leader.
\item When the \texttt{clientRequestResponse} function is called with
    a successful status, the original callback argument of the
    \texttt{clientRequest} function is called with the results from
    the RPC response.
\end{enumerate}

@@

The chat application extends the basic Raft.js over WebRTC client. The
chat application provides a large text area containing the chat
history and a text entry field below it
(see Figure~\ref{fig:browser_chat_1}). When the user clicks on the
\emph{Send} button (or presses enter), the current node ID is
prepended to the beginning of the text line and it is added to
a pending send queue and the text entry field is blanked. A function
to flush sends runs periodically (currently every 100ms) and does the
following:
% TODO: 100ms above should be less

\begin{enumerate}
\item If there is line of text that has been sent but not yet
    added (applied) to the state machine or there are no lines of text
    on the pending send queue, then the function does nothing.
\item A text line is removed from the head of the pending send queue
    and stored in \emph{curSend} to indicate that a send is in
    progress. Then a client request is made using an atomic
    \emph{seqAppend} command to add the new text to the state machine
    (\emph{seqAppend} is described below).  The \emph{seqAppend}
    operation attempts to append the text line to a list at the
    \emph{history} key in the state machine.
\item When the atomic \emph{seqAppend} operation completes, it may
    have failed (if another text line was appended first), in which
    case the flush function puts the failed text line back on the
    pending send queue. Whether the send was successful or failed, the
    \emph{curSend} variable is set to nil/null to indicate that no
    send is in progress. The next time the flush function runs it will
    attempt to send again. % TODO: reword
\end{enumerate}

@@

The chat application also extends the default set of state machine
operations to add a \emph{seqAppend} operation. This is an atomic
operation that is implemented as follows:

\begin{enumerate}
\item the \emph{seqAppend} operation operates on a specially structured
    key/value pair in the state machine map. The value of the pair is
    another associative map that contains a counter and list structure
    where the actual chat lines are accumulated. \emph{seqAppend}
    operations are sent with both a count and then new value to
    append.
\item when a \emph{seqAppend} operation is applied to the state
    machine and the count value in the command matches the counter in
    the state machine's value, then the new line in the request is
    added to the value list and the counter is incremented. If the count
    value does not match the counter then the the operation is not
    applied. In either case, a list is returned in the
    \texttt{clientRequestResponse} that contains two values: a boolean
    indicating whether the \emph{seqAppend} operation succeeded and
    a number indicating the current value of the counter.
\item the first time the \emph{seqAppend} operation is used on a non-existent
    key, the structure described above is created with the counter set
    to 0 and the list structure is created with the single value that
    was passed in the \emph{seqAppend} operation.
\item clients that call the \emph{seqAppend} operation should check the
    status value in the result list. If the status is false, then the
    client should retry the operation. Regardless of success of
    failure, then client should store the current value of the counter
    for the next \emph{seqAppend} call or retry.
\end{enumerate}

\section{Results}

@@

To simplify the execution of the tests and to make the testing
infrastructure more portable, a docker container environment was used
to encapsulate and execute the SlimerJS environment. The container is
built using \emph{test/Dockerfile} which is in turn derived from
the ``fentas/slimerjs" docker image \cite{fentas:slimerjs}.
This custom Dockerfile was created because the upstream version of the
Dockerfile used an older version of XULRunner that did not support
more than 7 WebRTC clients communicating together simultaneously. This
issue was discovered during testing.

Writing tests using SlimerJS involves creating JavaScript programs
that are run within the SlimerJS environment. These scripts can create
browser instances (essentially headless browser tabs), query the status
of those pages and execute JavaScript within the context of the pages
themselves.

Three different tests were created to test the basic Raft over WebRTC
application and the chat application. Each of the tests takes
a parameter indicating the number of nodes/pages to launch and test
again. The test script files are:

\begin{enumerate}
    \item \emph{test/common.js}: a module of common functions used by
        the other tests scripts.
    \item \emph{test/test\_up.js}: launch and start an initial leader
        node/page, then start the rest of the nodes (if any) so that
        the leader will configure them into the cluster. Then test
        then waits for every node to be recorded (applied) into the log
        of every other node of the cluster. The test reports the
        number of milliseconds for the cluster to reach this state.
    \item \emph{test/test\_kill\_nodes.js}: similar to
        \emph{test/test\_up.js} but once all the nodes are configured
        into the cluster, this script then kills the leader plus
        a configurable number of nodes. By default the script kills
        just under half the nodes of the cluster. The test then waits
        for a new leader to be elected and remove the dead nodes from
        the cluster. The test reports both the number of milliseconds
        for the cluster to come up and the recover from the failure.
    \item \emph{test/test\_chat\_propagate.js}: similar to
        \emph{test/test\_up.js} but once the cluster is up, the test
        connects to one of the non-leader nodes and and sends a chat
        text line. The test then waits for the chat message to
        propagate to every other node's state machine. The test
        reports both the number of milliseconds for the cluster to
        come up and for the messages to propagate to every node.
\end{enumerate}

The tests were all run on a system with the following configuration:

\begin{enumerate}
    \item \emph{CPUs}: 8 Intel Core i7 processors running at 3.20GHz
    \item \emph{Memory}: 12 GB
    \item \emph{OS}: Ubuntu 14.04.1 LTS
    \item \emph{Docker}: version 1.3.1
    \item \emph{Docker image}:
    \begin{enumerate}
        \item \emph{OS}: Ubuntu 12.04.5 LTS
        \item \emph{SlimerJS}: version 0.9.6
        \item \emph{XULRunner}: 38.0.5
    \end{enumerate}
\end{enumerate}

Three tests were run at 10 different cluster sizes (odd cluster sizes
from 3 to 21 nodes): \emph{test/test\_up.js} and
\emph{test/test\_kill\_nodes.js} in both modes (kill only leader and kill just
under half the nodes).
Each of these tests was run with three different Raft election
timeout values: 100ms, 500ms, 1000ms. The
\emph{test/test\_chat\_propagate.js} test was run at
11 different cluster sizes (odd cluster sizes from 1 to 21 nodes).
Each test mode and cluster size was run 7 times for a total of 707
test runs (3*10*7*3 + 11*7)

@@

Figure~\ref{fig:propagate1} shows the results of running the
\emph{test/wait\_chat\_propagate.js} test to show the amount of time
that a stable cluster takes to propagate, commit and apply a single
chat message to the state machine on every cluster node using the
atomic \emph{seqAppend} command.  Figure~\ref{fig:propagate1} is
a scatterplot with each test run as a blue ``x" and a black line
representing the average at each cluster size.

\begin{figure}[!t]
    \begin{tikzpicture}
    \begin{axis}[%
        xlabel={Nodes}, ylabel={Time (seconds)},
        %xtick=data,
        xmin=0, ymin=0,
        grid=major,
        scatter/classes={%
            propagate_1_et1000={mark=x,draw=blue},
            propagate_1_et1000_average={mark=}}]
    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_1.dat};
    \addplot[scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_1_average.dat};
    \end{axis}
    \end{tikzpicture}
\caption{Propagate$,$ Commit$,$ and Apply Command}
  \label{fig:propagate1}
\end{figure}

Figure~\ref{fig:propagate100} is similar to
Figure~\ref{fig:propagate1} but shows the results of adding 100 chat
messages instead of just 1. The commands are serialized which means
that the previous command must be committed before the current command
can be committed. The test is complete once the last chat message
command has been both committed and applied to every node of the
cluster. Figure~\ref{fig:propagate1} is a scatterplot with each test
run as a red ``x" and a black line representing the average at each
cluster size.

\begin{figure}[!t]
    \begin{tikzpicture}
    \begin{axis}[%
        xlabel={Nodes}, ylabel={Time (seconds)},
        %xtick=data,
        xmin=0, ymin=0,
        grid=major,
        scatter/classes={%
            propagate_100_et1000={mark=x,draw=red},
            propagate_100_et1000_average={mark=}}]
    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_100.dat};
    \addplot[scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_100_average.dat};
    \end{axis}
    \end{tikzpicture}
\caption{Propagate$,$ Commit$,$ and Apply Command}
  \label{fig:propagate100}
\end{figure}

>>>>

In order to keep the replicated transaction log from growing
indefinitely, Raft nodes should periodically compact their logs. There
are many different ways to accomplish log compaction and the best way
will depend on application requirements and on the specific nature of
the distributed state machine. For example, if the state machine is
simply a single shared counter, then any log entries before the most
recently committed entry may be safely discarded.
The simplest and most generic solution is often to provide some way of
snapshotting the entire state machine (including the term and log
index it represents) to disk at which point all previously applied log
entries can be discarded. Snapshotting the state machine also means
that the implementation must provide a way for the leader to serialize
and send the current state machine to other nodes if their transaction
log is too old (e.g. when a new node is added)
\cite[Section~5]{raft_thesis:ongaro14}.

@@

The Raft protocol also provides strict linearizable semantics for
client commands/requests (reads and writes) for applications that
require it.
In order to accomplish this, each client is given a unique ID, and the
client assigns a unique serial number to each command.  This prevents
a single client command from being applied twice (leader crash,
network duplication) or lost (packet loss)
\cite[Section~6.3]{raft_thesis:ongaro14}.

@@

@@

\subsubsection{WebRTC Protocols}

@@

The protocols that make up the WebRTC suite are categorized into five
general functionality groups:
\ifdefined\OPTIONAL

\begin{enumerate}
    \item data transport
        \\
        \cite{draft-rtcweb-transports}
        \cite{draft-rtcweb-qos}
        \cite{RFC4960:sctp}
        \cite{RFC3758:sctp-reliability}
        \cite{draft-sctp-dtls-encaps}
        \cite{draft-mmusic-sctp-sdp}
        \cite{draft-sctp-ndata}
        \cite{draft-rtcweb-rtp-usage}
        \cite{draft-rtcweb-data-channel}
        \cite{draft-rtcweb-data-protocol}
        \cite{draft-rtcweb-alpn}
        \cite{RFC5245:ice}
        \cite{RFC5128:p2p-nat}
        \cite{RFC5389:stun}
        \cite{RFC3489:stun-udp}
        \cite{RFC5766:turn}
        \cite{RFC6156:turn-ipv6}
        \cite{RFC6544:tcp-ice}
        \cite{RFC4571:rtp-rtcp}
        \cite{RFC5764:dtls-srtp}
    \item data framing and security
        \\
        \cite{RFC3550:rtp}
        \cite{RFC3711:srtp}
        \cite{draft-rtcweb-rtp-usage}
        \cite{draft-rtcweb-data-channel}
        \cite{draft-rtcweb-data-protocol}
    \item connection management
        \\
        \cite{RFC2327:sdp}
        \cite{RFC3264:sdp-offer-answer}
        \cite{draft-rtcweb-jsep}
        \cite{RFC5245:ice}
        \cite{draft-mmusic-trickle-ice}
        \cite{RFC5763:srtp-using-dtls}
    \item data formats
    \item presentation and control
    \item local system support functions
\end{enumerate}

The last three functionality groups focus on interactive audio and video
applications and are less relevant to Raft over WebRTC.
Some of the important standards for the first three functionality
groups are cited above.
\else
data transport, data framing and security, connection management, data
formats, presentation and control, local system support functions.
The last three functionality groups focus on interactive audio and
video applications and are less relevant to Raft over WebRTC.
\fi

@@

\section{Design and Implementation}

\begin{figure}[!t]
  \centerline{\includegraphics[width=0.45\textwidth]{imgs/raft_rtc_architecture}}
  \caption{Raft over RTC Architecture}
    \label{fig:raft_rtc_architecture}
\end{figure}

\subsection{Raft.js Core Library}

The modular design of the Raft.js implementation allows it to be
easily used in different contexts. For example, the \texttt{RaftServerLocal}
class (\texttt{local.js}) implements RPC calls using plain JavaScript function
calls between different instances of the class (nodes) in the same
execution context. For example, the online Raft visualization at
http://kanaka.github.io/raft.js/
%% \href{http://kanaka.github.io/raft.js/}{http://kanaka.github.io/raft.js/}
\cite{raft.js:visualization} uses the \texttt{RaftServerLocal} class
and instantiates it using a scheduling function that executes one
scheduled event each time the user clicks on the ``Step" button rather
than due to the passage of time.

The current version of Raft.js implements most of the Raft algorithm
including: leader election, log replication, safety, membership
changes, basic client interaction, and read-only operation
optimizations. % TODO: read-only operations needs mentioning earlier
Log compaction and full client linearizability are not currently
implemented (clients and client transactions are not assigned unique
IDs). Log compaction is important for long-running production systems
in order to limit memory growth and allow new nodes to catch up their
logs more efficiently. However log compaction is not required for
verifying overall system correctness and does not significantly impact
short-running test configurations.

Full client linearizability is an important feature for simplifying
certain types of distributed applications, however the implementation
of linearizability requires some significant additions to the Raft
implementation (e.g. assigning client IDs and sequence IDs and
tracking timed-out clients). For the chat application, simply adding
a single atomic check-and-set state machine command is sufficient to
achieve a linearized chat history.  This does impose a slight
additional cost on the client because it must retry the atomic
operations until they succeed. The author intends to implement both
log compaction and full client linearizability in the future, however
these do not have a significant impact on the ability to test Raft
combined with WebRTC.

\begin{enumerate}
\item The paper describes the Raft RPCs in terms of synchronous stream
    based request/response model where requests and responses are
    automatically correlated together by a lower level of the software
    stack. For example with HTTP 1.0 \cite{http:rfc1945} client
    requests are followed by a server response on any given TCP
    stream. Most remote procedure call mechanisms also use
    a synchronous request/response model. %TODO/CITE
    WebRTC DataChannel connections are message based rather than
    request/response stream based. %TODO/CITE
    In the original Raft definition of RPC, request and response messages
    almost contain enough information to function correctly in an
    environment without builtin request/response correlation, such as
    UDP or SCTP when used with reliable transport disabled. %TODO/CITE
    Raft.js includes a small modification to the response RPCs for
    requestVote and appendEntries to add a `sourceId' field that
    enables the protocol to function with message based transports.
\item The Raft paper is clear that each membership change must be
    delayed until all the previous membership change is committed
    (sequential) \cite[Section~4.1]{raft_thesis:ongaro14}. It is
    implied that the leader node should track all pending membership
    change requests and apply them one at a time.  However, in
    Raft.js, if a client requests a membership change while another
    change request is pending (uncommitted), the leader will reject
    the change request with a status of `PENDING\_CONFIG\_CHANGE'.
    This keeps the core Raft implementation simpler with the trade-off
    being that the client implementation of membership change requests
    is slightly more complex due to retry logic.
\end{enumerate}

@@

The file `rtc\_server.js' implements the signaling server for the Raft
over WebRTC application. The server extends the PeerJS
`ExpressPeerServer' object in order to provide WebRTC signaling. In
addition to the WebRTC signaling function the server also provides the
following service endpoints:

@@

Subsequent nodes are added to the same Raft cluster by starting a new
browser context (separate browser, browser window, or browser tab) and
loading the same URL without the ``firstServer" fragment identifier
(Figure~\ref{fig:raft_rtc_sequence}~[9]). There is a convenience link
provided in the application page that opens a new browser window with
a new cluster node.  %(see \ref{fig:browser_1}).

@@

\subsubsection{Client Statistics}

In addition to the low level Raft message log, the following
statistics are shown in the application:
\ifdefined\OPTIONAL

\begin{enumerate}
\item \emph{Term}: the current Raft leadership term.
\item \emph{State}: the current Raft node state (follower, candidate,
    or leader)
\item \emph{Cluster Size}: a count of the current Raft cluster
    membership (number of nodes currently participating in the
    cluster)
\item \emph{Log Length}: number of entries in the transaction log.
\item \emph{Request Votes Sent/Received}: the number of `requestVote`
    RPCs that have been sent from and received by this node.
\item \emph{Append Entries Sent/Received}: the number of
    `appendEntries' RPCs that have been sent from and received by this
    node.
\end{enumerate}
\else
the current Raft leadership term, the current Raft node state
(follower, candidate, or leader), Raft cluster size (number of
participating nodes), number of entries in the transaction log, and
number of request vote and append entries RPCs that have been sent and
received.
\fi

@@

In the Raft over WebRTC implementation each local browser is a client
of the Raft cluster. However in Raft, the leader node is the only node
that adds entries to the log so clients on follower nodes must have
a way to communicate with the leader.
\ifdefined\OPTIONAL
This is implemented as a pair of
functions named \texttt{clientRequest} and
\texttt{clientRequestResponse}. The \texttt{clientRequest} function is
called by the client with the arguments of a client request RPC and
a callback function to be called once the request is completed. The
\texttt{clientRequest}/\texttt{clientRequestResponse} functions work
as follows:

\begin{enumerate}
\item The \texttt{clientRequestResponse} is registered with the local
    Raft.js implementation as the callback function anytime
    a \texttt{clientRequestResponse} RPC is received.
\item The \texttt{clientRequest} function keeps track of the current
    leader ID.  If the current leader ID is not set or this node is
    the leader then the client request RPC is sent directly to the
    local Raft node bypassing the network transport layer.  If the
    current leader ID is set then the request is sent to that node as
    a normal RPC request over the WebRTC channel.
\item When the \texttt{clientRequestResponse} function is called and
    the status is ``NOT\_LEADER" then the current leader ID being
    tracked by the \texttt{clientRequest} function is set to the
    leader ID specified in the response. Then the original request is
    re-issued over the normal WebRTC channel to the leader.
\item When the \texttt{clientRequestResponse} function is called with
    a successful status, the original callback argument of the
    \texttt{clientRequest} function is called with the results from
    the RPC response.
\end{enumerate}
\else
When a \texttt{clientRequest} message is sent to a local Raft node
that is not the leader, the \texttt{clientRequestResponse} message
will be returned with a status of ``NOT\_LEADER" and the ID of the
current leader. The \texttt{clientRequest} message will then be
re-issued over the WebRTC DataCannel to the leader ID that was
returned.
\fi

@@

\begin{figure}[!t]
  \centerline{\includegraphics[width=0.45\textwidth]{imgs/raft_rtc_sequence}}
  \caption{Raft over RTC Sequence Diagram}
    \label{fig:raft_rtc_sequence}
\end{figure}

The chat application extends the basic Raft.js over WebRTC client. The
chat application provides a large text area containing the chat
history and a text entry field below it. % (see Figure~\ref{fig:browser_chat_1}).
When the user clicks on the \emph{Send} button (or presses enter), the
current node ID is prepended to the beginning of the text line and it
is added to a pending send queue and the text entry field is blanked.
\ifdefined\OPTIONAL
A function to flush the send queue runs periodically (currently every
100ms) and does the following:
% TODO: 100ms above should be less

\begin{enumerate}
\item If there is line of text that has been sent but not yet
    added (applied) to the state machine or there are no lines of text
    on the pending send queue, then the function does nothing.
\item A text line is removed from the head of the pending send queue
    and stored in \emph{curSend} to indicate that a send is in
    progress. Then a client request is made using an atomic
    \emph{seqAppend} command to add the new text to the state machine
    (\emph{seqAppend} is described below).  The \emph{seqAppend}
    operation attempts to append the text line to a list at the
    \emph{history} key in the state machine.
\item When the atomic \emph{seqAppend} operation completes, it may
    have failed (if another text line was appended first), in which
    case the flush function puts the failed text line back on the
    pending send queue. Whether the send was successful or failed, the
    \emph{curSend} variable is set to nil/null to indicate that no
    send is in progress. The next time the flush function runs it will
    attempt to send again. % TODO: reword
\end{enumerate}
\else
A function to flush the send queue runs periodically (currently every
100ms) to send the chat message at the head of the queue using
the \emph{seqAppend} operation.
\fi

@@

The chat application extends the default set of state machine
operations to add the \emph{seqAppend} operation. This is an atomic
operation that is implemented as follows:

\begin{enumerate}
\item The \emph{seqAppend} operation operates on a specially
    structured key/value pair in the state machine map. The value of
    the pair is another associative map that contains a counter and
    list structure where the actual chat lines are accumulated.
    \emph{seqAppend} operations are sent with both a count and the new
    value to append.
\item When a \emph{seqAppend} operation is applied to the state
    machine and the count value in the command matches the counter in
    the state machine's value, then the new line in the request is
    added to the value list and the counter is incremented. If the
    count value does not match the counter then the the operation is
    not applied. In either case, a list is returned in the
    \texttt{clientRequestResponse} that contains two values: a Boolean
    indicating whether the \emph{seqAppend} operation succeeded and
    a number indicating the current value of the counter.
\item The first time the \emph{seqAppend} operation is used on
    a non-existent key, the structure previously described is created
    with the counter set to 0 and the list structure is created with
    the single value that was passed in the \emph{seqAppend}
    operation.
\item Clients that call the \emph{seqAppend} operation should check
    the status value in the result list. If the status is false, then
    the client should retry the operation. Regardless of success of
    failure, then client should store the current value of the counter
    for the next \emph{seqAppend} call or retry.
\end{enumerate}

\section{Experimental Results}

@@

To simplify the execution of the tests and to make the testing
infrastructure more portable, a docker container environment
encapsulates and executes the SlimerJS environment. The container is
built using \emph{test/Dockerfile} which is in turn derived from the
``fentas/slimerjs" docker image \cite{fentas:slimerjs}.
This custom Dockerfile was created because the upstream version of the
Dockerfile used an older version of XULRunner that did not support
more than 7 WebRTC clients communicating together simultaneously. This
issue was discovered during initial testing.

Writing tests using SlimerJS involves creating JavaScript test
programs which start a \emph{Twst} server instance and then launch a SlimerJS
docker container for each node of the cluster. Each container starts
a XULRunner browser instance of the Raft over WebRTC chat application.
Also included with the application is the \emph{Twst} client library
which connects back to the \emph{Twst} service in the test program.
Once all the chat application nodes have connected to the test
server/program, then the actual test is started across all the
application nodes.

Three different tests programs/servers were created to test the basic
Raft over WebRTC application. Each of the tests takes a parameter
indicating the number of nodes/pages to launch. The test source files
are:

\begin{enumerate}
    \item \emph{test/common.js}: a module of common functions used by
        the other tests scripts.
    \item \emph{test/test\_up.js}: launch and start an initial leader
        node/page, then start the rest of the nodes (if any) so that
        the leader will configure them into the cluster. Then test
        then waits for every node to be recorded (applied) into the
        log of every other node of the cluster. The test reports the
        number of milliseconds for the cluster to reach this state.
    \item \emph{test/test\_kill\_nodes.js}: similar to
        \emph{test/test\_up.js} but once all the nodes are configured
        into the cluster, this script then kills the leader plus
        a configurable number of nodes. By default the script kills
        just under half the nodes of the cluster. The test then waits
        for a new leader to be elected and remove the dead nodes from
        the cluster. The test reports both the number of milliseconds
        for the cluster to come up and the number of milliseconds for
        the cluster to fully recover from the failure and evict the
        failed nodes.
    \item \emph{test/test\_chat\_propagate.js}: similar to
        \emph{test/test\_up.js} but once the cluster is up, the test
        connects to one of the non-leader nodes and and sends a chat
        text line. The test then waits for the chat message to
        propagate to every other node's state machine. The test
        reports both the number of milliseconds for the cluster to
        come up and for the messages to propagate to every node.
\end{enumerate}

\ifdefined\OPTIONAL
The tests were all run on a system with the following configuration:

\begin{enumerate}
    \item \emph{CPUs}: 8 Intel Core i7 processors running at 3.20GHz
    \item \emph{Memory}: 12 GB
    \item \emph{OS}: Ubuntu 14.04.1 LTS
    \item \emph{Docker}: version 1.3.1
    \item \emph{Docker image}:
    \begin{enumerate}
        \item \emph{OS}: Ubuntu 12.04.5 LTS
        \item \emph{SlimerJS}: version 0.9.6
        \item \emph{XULRunner}: 38.0.5
    \end{enumerate}
\end{enumerate}
\else
The tests were all run on a system with 8 Intel Core i7 processors
running at 3.20GHz and 12 GB of main memory. The Docker image used to
run the test was built using Ubuntu 12.04.5 LTS with \emph{SlimerJS}
version 0.9.6, and \emph{XULRunner} version 38.0.5.
\fi

Five different tests (using different options for the three test
programs) were run at 17 different cluster sizes (odd cluster sizes
from 3 to 35 nodes). Each of these tests was run with three different
Raft election timeout values: 250ms, 500ms, 1000ms. Each of the five
tests and cluster size combinations was run 7 times.

@@

\begin{figure}[!t]
    \begin{tikzpicture}
    \begin{axis}[%
        xlabel={Nodes}, ylabel={Time (seconds)},
        %xtick=data,
        xmin=0, ymin=0,
        grid=major,
        scatter/classes={%
            propagate_1_et1000={mark=x,draw=blue},
            propagate_1_et1000_average={mark=}}]
    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_1.dat};
    \addplot[scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_1_average.dat};
    \end{axis}
    \end{tikzpicture}
    \caption{Propagate$,$ Commit$,$ and Apply 1 Command}
    \label{fig:propagate1}
%
    \vspace{1cm}
%
    \begin{tikzpicture}
    \begin{axis}[%
        xlabel={Nodes}, ylabel={Time (seconds)},
        %xtick=data,
        xmin=0, ymin=0,
        grid=major,
        scatter/classes={%
            propagate_100_et1000={mark=x,draw=blue},
            propagate_100_et1000_average={mark=}}]
    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_100.dat};
    \addplot[scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_100_average.dat};
    \end{axis}
    \end{tikzpicture}
    \caption{Propagate$,$ Commit$,$ and Apply 100 Commands}
    \label{fig:propagate100}
\end{figure}
